---
title: "sisg-aqg-2015-p3"
author: "SISG module 23"
date: "July 24, 2015"
output: html_document
---


```{r}
# ===============================================================
# Summer Institute in Statistical Genetics, 2015, Seattle
# Advanced Quantitative Genetics
# Practical 3 
# Date:
# Author:
# ===============================================================
```


```{r}
library(knitr)
# see https://github.com/yihui/knitr/issues/277
opts_knit$set(root.dir ='C:/Users/vonholle/Documents/grad_school/misc/training/sisg-2015/module23/sisg-2015-module23')
#setwd("C:/Users/vonholle/Documents/grad_school/misc/training/sisg-2015/module23/sisg-2015-module23") # Windows

#opts_chunk$set(cache = TRUE, cache.path = "knitrcache/test-")
opts_chunk$set(cache = TRUE)

#install.packages("Haplin")
library(Haplin)
#install.packages("qqman")
library(qqman)
```

```{r}
alpha = 5e-8
threshold = qchisq(1-alpha, 1)
q2 = 0.005
n = 10000
ncp = n*q2/(1-q2)
power = 1-pchisq(threshold, 1, ncp)
threshold
ncp
power
```

#### Note: for following runs in plink2 may want to double check in plink because plink2 is in beta development.

### Exercise 1

* How many individuals would you need to detect a locus that explains 1\% with a power of 80\%?


## Let's attempt to use PLINK to convert the .ped and .map files to binary format.

```{r}
# Once you've set up your binary directory for this analysis and have set R to that directory, do this:

# Convert from .ped and .map PLINK format to binary
system("plink2 --bfile practical_3/data/QIMRX --make-bed --out QIMRX")
```


## Running a GWAS

Given this short overview of PLINK we will now attempt to perform a GWAS on data from human height and serum transferrin levels. This will be done in three steps:

* File inspection
* QC
* GWAS for human height and serum transferrin level


### File inspection
Firstly, we will use R to inspect the .fam, .bim, and HT_T_X.pheno files. Let's have a quick look around these files and establish some of the properties of the data using R. As always, let's firstly read in the data.

```{r}
fam <- read.table("practical_3/data/QIMRX.fam")
bim <- read.table("practical_3/data/QIMRX.bim")
pheno <- read.table("practical_3/data/HT_T_X.pheno")
```

```{r}
# Count the number of genotyped individuals. Number of rows
dim(fam)
head(fam)
```

```{r}
# Count the number of genotyped SNPs
dim(bim)
head(bim)
```

```{r}
# Count the number of individuals with height and transferrin
# measurements
dim(pheno)
dim(pheno[complete.cases(pheno), ])
head(pheno)
```


### Quality Control

The quality control steps are critical to any analysis and often take more time then the analyses themselves. We will now work with PLINK 1.9 to do some QC on the binary files for the QIMR data on human height and transferrin levels. These QC step will include

* Estimate the allele frequencies for all SNPs

* Calculate SNP and individual missingness

* Calculate p-values for Hardy-Weinberg (HW)

Each of the below commands should produce a le with an extension that is indicative of the process used. Unlike previous programs PLINK allows for an out directory to be specified with --out. We will put all our results in the practical 3/results folder. Execute the following command line arguments to produce three files for allele frequency, missingness, and HW.

```{r}
system("plink2 --bfile practical_3/data/QIMRX --freq --out practical_3/results/QIMRX", intern = F)
system("plink2 --bfile practical_3/data/QIMRX --missing --out practical_3/results/QIMRX", intern = F)
system("plink2 --bfile practical_3/data/QIMRX --hardy --out practical_3/results/QIMRX", intern = F)

```

Mini Exercise - Run these same commands but with PLINK 1 to recognise the incredible differences in speed that PLINK 2 others. For large analyses these speed ups make all the difference between being able to complete analyses or not { additionally, RAM is often the limiting factor rather than CPU time and PLINK v1.9 has a much lower memory profile when compared to the first version.

Read the resultant files into R and attempt to answer the following questions

* How many SNPs have MAF > 0.05?

* How many individuals have missingness > 10%?

* How many SNPs have missingness of > 1%?

* How many SNPs have a HWE p-value < 0.001?

Below are some hints on how to do this

```{r}
# Read in the .frq file
frq <- read.table("practical_3/results/QIMRX.frq", header = T, na = "NA")
head(frq)
```

```{r}
# Find the proportion of those with missingness less than 0.05
prop.lw.maf <- sum((frq$MAF < 0.05)) / length(frq$MAF)
# The sum component above is summing up the true values
# Look at what (frq$MAF < 0.05) does to get a better feel for
# this solution. Can be done in many other ways
prop.lw.maf
# Do something similar for the rest of the questions
```


### Running a GWAS
The work above was designed to investigate what PLINK can do on-the-fly with the filtering commands in listing 7. We can do all of these steps and the GWAS in one command with PLINK. In the terminal execute a similar command as in listing 7 using the quality control flags from listing 3 along with the --assoc flag to run a GWAS with MAF filter 0.05, individual missing rate 0.1, SNP missing rate 0.01, and Hardy-Weinberg of 0.001. Note that this will be quite a long terminal command. Remember to give the out path to the practical 3/results folder.


```{r}

system("plink2 --bfile practical_3/data/QIMRX --maf 0.05 --mind 0.1 --geno 0.01 --hwe 0.001 --pheno practical_3/data/HT_T_X.pheno --assoc --out practical_3/results/gwas_pheno_1", intern = F)

```


Once the files have been moved, read the association results into R. We will first draw a manhattan plot.

```{r}
# Read in the association results
gwas.res <- read.table("practical_3/results/gwas_pheno_1.qassoc",
header = T)
dim(gwas.res)
head(gwas.res)
```

```{r}
# Build the data frame that the manhattan plot function requires
man.df <- data.frame(gwas.res$BP, gwas.res$CHR, gwas.res$P, gwas.res$SNP)
# Load the manhattan plot library. If not installed use install.packages("qqman")
#library(qqman)
# Rename the columns for the manhattan function
colnames(man.df) <- c("BP", "CHR", "P", "SNP")
# Produce the manhattan plot. NOTE THAT THIS MAY TAKE SOME TIME AND MAY CRASH YOUR COMPUTER
# IF IT HAS POOR RESOURCES
manhattan(man.df)
# Drawing a qq plot
# -------------------
obs.p <- gwas.res$P
# Order these
obs.p.srt <- sort(obs.p)
m.log10.obs.p <- -(log10(obs.p.srt))
max.p <- max(m.log10.obs.p)
exp.val <- seq(1, length(obs.p))
m.log10.exp.val <- -log10((exp.val - 0.5) / length(exp.val))
plot(c(0, max.p), c(0, max.p), col = "red", lwd = 2, type = "l",
xlab = "Expected -log10(p)", ylab = "Observed -log10(p)",
xlim = c(0, max.p), ylim = c(0, max.p), las = 1, xaxs = "i",
yaxs = "i", bty = "l", main = "Trait 1")
points(m.log10.exp.val, m.log10.obs.p)
# Alternatively you can use a package. This has the added component of
# having confidence interval bounds

#install.packages("Haplin")
#library(Haplin)
x <- pQQ(obs.p.srt, nlabs = 6, conf = 0.95)
```

## Exercise 2

* Repeat the GWAS for the second phenotype.
  * Note to myself: Be sure to rename the output file for the second phenotype and specify which phenotype using the \textrm{--mpheno} flag (followed by ''2`` in this case). By default the first column in the .pheno file is used.
  
```{r}
system("plink2 --bfile practical_3/data/QIMRX --maf 0.05 --mind 0.1 --geno 0.01 --hwe 0.001 --pheno practical_3/data/HT_T_X.pheno --mpheno 2 --linear --out practical_3/results/gwas_pheno_2", 
intern = F)
```


* Produce the same plots as for height

#### Note to myself: You have to rename the qassoc.linear file to .qassoc only to make the read.table function work.

```{r}
# first need to rename file because it won't work with names following 2 periods.
file.rename(from = file.path("practical_3/results/", "gwas_pheno_2.qassoc.linear"), 
            to = file.path("practical_3/results/", "gwas_pheno_2.qassoc"))

# Read in the association results
gwas.res2 <- read.table("practical_3/results/gwas_pheno_2.qassoc",
header = T)
dim(gwas.res2)
head(gwas.res2)
```

#### Note: this part takes a very long time.
```{r}
# Build the data frame that the manhattan plot function requires
man.df <- with(gwas.res2, data.frame(BP, CHR, P, SNP))
# Load the manhattan plot library. If not installed use install.packages("qqman")
#library(qqman)
# Rename the columns for the manhattan function
colnames(man.df) <- c("BP", "CHR", "P", "SNP")
# Produce the manhattan plot. NOTE THAT THIS MAY TAKE SOME TIME AND MAY CRASH YOUR COMPUTER
# IF IT HAS POOR RESOURCES
manhattan(man.df)
# Drawing a qq plot
# -------------------
obs.p <- gwas.res2$P
# Order these
obs.p.srt <- sort(obs.p)
m.log10.obs.p <- -(log10(obs.p.srt))
max.p <- max(m.log10.obs.p)
exp.val <- seq(1, length(obs.p))
m.log10.exp.val <- -log10((exp.val - 0.5) / length(exp.val))
plot(c(0, max.p), c(0, max.p), col = "red", lwd = 2, type = "l",
xlab = "Expected -log10(p)", ylab = "Observed -log10(p)",
xlim = c(0, max.p), ylim = c(0, max.p), las = 1, xaxs = "i",
yaxs = "i", bty = "l", main = "Trait 1")
points(m.log10.exp.val, m.log10.obs.p)
# Alternatively you can use a package. This has the added component of
# having confidence interval bounds

#install.packages("Haplin")
#library(Haplin)
x <- pQQ(obs.p.srt, nlabs = 6, conf = 0.95)
```


* Calculate $\lambda_{GC}$ for both phenotypes.

#### Note: not sure how you get the lambda

## Exercise 3

* Calculate the first 10 principal components of the genotype matrix

* \textrm{plink2 --bfile practical 3/data/QIMRX --pca 10 --out practical_3/results/QIMRX}


```{r, eval=FALSE, echo=F}
system("plink2 --bfile practical_3/data/QIMRX --pca 10 --out practical_3/results/QIMRX", intern=F)
```


* Rerun the association analysis with these 10 PCs

```{r, eval=FALSE, echo=F}
system("plink2 --bfile practical_3/data/QIMRX --maf 0.05 --geno 0.1 --mind 0.01 -hwe 0.001 --linear --covar practical_3/results/QIMRX.eigenvec --pheno practical_3/data/HT_T_X.pheno --mpheno 2 --out practical_3/results/QIMRX_ST2",
       intern=F)
```

* Read the results back into R

#### Note to myself: You have to rename the qassoc.linear file to .qassoc only to make the read.table function work.


```{r}
# first need to rename file because it won't work with names following 2 periods.
file.rename(from = file.path("practical_3/results/", "QIMRX_ST2.assoc.linear"), 
            to = file.path("practical_3/results/", "QIMRX_ST2.assoc"))


# Read in the association results
gwas.res3 <- read.table("practical_3/results/QIMRX_ST2.assoc",
header = T)
dim(gwas.res3)
head(gwas.res3)

```


* Subset the data to only leave the estimates for the SNPs

* \textrm{man.df <- subset(gwas.res, TEST=='ADD',c('BP','CHR','P','SNP'))}

```{r}
# Build the data frame that the manhattan plot function requires
#man.df <- with(gwas.res3, data.frame(BP, CHR, P, SNP))
man.df <- subset(gwas.res3, TEST=='ADD', c('BP','CHR','P','SNP'))
```

* Alternatively you can use grep "ADD" over the output file.

* Re-draw the manhattan plots

```{r}
# Produce the manhattan plot. NOTE THAT THIS MAY TAKE SOME TIME AND MAY CRASH YOUR COMPUTER
# IF IT HAS POOR RESOURCES
manhattan(man.df)
# Drawing a qq plot
# -------------------
obs.p <- gwas.res3$P
# Order these
obs.p.srt <- sort(obs.p)
m.log10.obs.p <- -(log10(obs.p.srt))
max.p <- max(m.log10.obs.p)
exp.val <- seq(1, length(obs.p))
m.log10.exp.val <- -log10((exp.val - 0.5) / length(exp.val))
plot(c(0, max.p), c(0, max.p), col = "red", lwd = 2, type = "l",
xlab = "Expected -log10(p)", ylab = "Observed -log10(p)",
xlim = c(0, max.p), ylim = c(0, max.p), las = 1, xaxs = "i",
yaxs = "i", bty = "l", main = "Trait 1")
points(m.log10.exp.val, m.log10.obs.p)
# Alternatively you can use a package. This has the added component of
# having confidence interval bounds

#install.packages("Haplin")
#library(Haplin)
x <- pQQ(obs.p.srt, nlabs = 6, conf = 0.95)
```


* Calculate $\lambda_{GC}$ for both phenotypes.

* You may find the following command useful to clump your results into ``roughly'' independent regions
  * \textrm{plink --bfile practical\_3/data/QIMRX --clump QIMRX_ST2.assoc.linear --clump-p1 0.5 --clump-p2 0.5 --clump-r2 0.20 --clump-kb 500 --out hgt gwas clump}

## Additional exercise

Use PLINK to do a test for dominance for the top 5 SNPs for height. Is there any evidence for dominance?

```{r}
system("plink2 --bfile practical_3/data/QIMRX --extract practical_3/data/top_snps.txt --pheno practical_3/data/HT_T_X.pheno --mpheno 1 --linear genotypic  --out practical_3/results/dom_test", intern = F)
```

```{r}
# first need to rename file because it won't work with names following 2 periods.
file.rename(from = file.path("practical_3/results/", "dom_test.assoc.linear"), 
            to = file.path("practical_3/results/", "dom_test.assoc"))

# print off variance info
cat(readChar("practical_3/results/dom_test.assoc", 1e5))

```

